---
title: "Supplement to:  \nA Method to Standardize Application Scores"
author: "Joseph Policarpio, Colin Van Orman, MS, Nora Few, PhD, and Ruoqing Zhu, PhD"
date: "12/9/2020"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) # knitr::purl("Supplemental Digital Appendix.Rmd", documentation = 0)
```


# Background

Nearly every medical school admissions office is tasked with evaluating thousands of medical school applications per year – far too many for any individual reviewer to score. Earlier this year, we explored scaling, standardizing, and normalizing medical school application scores from different reviewers. We developed a model and simulated 100 applications each scored by three reviewers. We found that our method of transforming each set of a single reviewer’s scores into a Gaussian distribution before merging with sets of scores provided by other reviewers may provide value in highlighting applications with the highest, or lowest scores provided by a single reviewer. Our method may help admissions offices to reduce inter-reviewer cognitive biases and allow for a method of standardization beyond a typical scoring rubric.

In a preliminary literature review, Pubmed search terms "Search (((school[Title]) AND ((admissions[Title]) OR admission[Title])) AND scor\*[Title/Abstract]) AND ((standard\*[Title/Abstract]) OR normal\*[Title/Abstract])" yielded 21 results, none of which included discussion on reviewing or scoring processes, nor a normalization of such scores. Further Pubmed search terms "(secondary application[Title/Abstract]) OR supplemental application[Title/Abstract] AND admission*[Title/Abstract]" yielded no results that mentioned the selection process.

Here we propose a method of scaling, standardizing and normalizing scores from various reviewers.

# Simulate Three Reviewers: Harsh, Typical, and Lenient

We will first assume a panel of three reviewers of varying strictness. A harsh, who grades on a more difficult scale than the average reviewer and thus provides lower scores of applications, is considered here to be a Hawk. A reviewer who grades on a more lenient scale and thus provides higher scores is considered here to be a Dove. An average reviewer here is given the name Average Jane.

The scores provided by each of the reviewers in the panel will be modeled using the R programming language. In the first example, to demonstrate proof of concept, we will have each reviewer score 100 applications.

Assume the possible range of scores to be from 0 to 50.

First we load the required RStudio computing packages, then set N, the number of applications to 100.

```{r}
library(timeDate)
library(timeSeries)
library(fBasics)
library(fGarch)

N = 100

a = 175 # transparency value of histogram color, out of 255
```

We then model the scores provided by each of the three reviewers. The means and standard deviations are selected arbitrarily here.

Hawk (red): Right skewed distribution with mean of 9 and standard deviation of 4. The scores are then displayed on a histogram.
```{r}
set.seed(1)
hawk1 = rsnorm(N, mean = 9, sd = 4, xi = 3)
hawk = data.frame(Raw_Hawk = as.vector(round(hawk1)))
hawk$Raw_Hawk_Sorted = sort(hawk$Raw_Hawk)
ymax = 15 # max(hawk$Raw_Hawk)

hawk_hist = hist(hawk$Raw_Hawk_Sorted, breaks = seq(min(hawk$Raw_Hawk)-1,
  max(hawk$Raw_Hawk), by=1),  col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
  xlim=range(0,50), ylim=c(0,15), main="Hawk Scores - Raw", xlab="Score",
  ylab="Frequency")
```

Jane (violet): Normal distribution with mean of 19 and standard deviation of 7. The scores are then displayed on a histogram.
```{r}
set.seed(1)
jane1 = rnorm(N, 19, 7)
jane = data.frame(Raw_Jane = as.vector(round(jane1)))
jane$Raw_Jane_Sorted = sort(jane$Raw_Jane)

jane_hist = hist(jane$Raw_Jane_Sorted, breaks = seq(min(jane$Raw_Jane)-1,
  max(jane$Raw_Jane), by=1), col=rgb(0,0,255,a, maxColorValue = 255), lwd=2, 
  xlim=range(0,50), ylim=c(0, ymax), main="Jane Scores - Raw", xlab="Score", 
  ylab="Frequency")
```

Dove (green): Left skewed distribution with mean of 40 and standard deviation of 4. The scores are then displayed on a histogram.
```{r}
set.seed(1)
dove1 = rsnorm(N, mean = 40, sd = 4, xi = 0.5)
dove = data.frame(Raw_Dove = as.vector(round(dove1)))
dove$Raw_Dove_Sorted = sort(dove$Raw_Dove)

dove_hist = hist(dove$Raw_Dove_Sorted, breaks = seq(min(dove$Raw_Dove)-1, 
  max(dove$Raw_Dove), by=1), col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
  xlim=range(0,50), ylim=c(0,ymax), main="Dove Scores - Raw", xlab="Score",
  ylab="Frequency")
```

A table showing the first 20 application scores is displayed.
```{r}
mydata = data.frame(App_No = c(1:N),
                    Hawk_raw = hawk$Raw_Hawk_Sorted,
                    Jane_raw = jane$Raw_Jane_Sorted,
                    Dove_raw = dove$Raw_Dove_Sorted
                                          )
head(mydata, n=20)

```

All three histograms are then displayed together. The kernel density of all scores is shown with the black line.
```{r}
ymax = max(c(hawk_hist$counts, jane_hist$counts, dove_hist$counts))

hist(hawk$Raw_Hawk_Sorted, breaks = seq(min(hawk$Raw_Hawk)-1, max(hawk$Raw_Hawk), by=1),
  col=rgb(255,0,0,a, maxColorValue = 255), lwd=2, xlim=range(0,50), 
  ylim=range(0,ymax), main=NULL, xlab=NULL, ylab=NULL)
par(new=TRUE)
hist(jane$Raw_Jane_Sorted, breaks = seq(min(jane$Raw_Jane)-1, max(jane$Raw_Jane), by=1),
  col=rgb(0,0,255,a, maxColorValue = 255), lwd=2, xlim=range(0,50),
  ylim=range(0,ymax), main=NULL, xlab=NULL, ylab=NULL)
par(new=TRUE)
hist(dove$Raw_Dove_Sorted, breaks = seq(min(dove$Raw_Dove)-1, max(dove$Raw_Dove), by=1),
  col=rgb(0,255,0,a, maxColorValue = 255), lwd=2, xlim=range(0,50),
  ylim=range(0,ymax), main="Raw Scores", xlab="Score", ylab="Frequency")

scale_factor = N*3
x = seq(0, 40, length.out = 3*N)
all3 = sort(c(hawk$Raw_Hawk, jane$Raw_Jane, dove$Raw_Dove))

y2 = density(all3)
y5 = scale_factor*y2$y
lines(y2$x, y5, col = "black", lwd = 2)
```

# Applying our Method f(x) = Q(x/(n+1))

The scores are first ranked from 1 to the total number of applications reviewed, then transformed to a normal distribution using the quantile function and x/(n+1) copula.
```{r}
hawk$rank = rank(hawk$Raw_Hawk)
jane$rank = rank(jane$Raw_Jane)
dove$rank = rank(dove$Raw_Dove)

hawk$normal = qnorm(hawk$rank/(1+N))
jane$normal = qnorm(jane$rank/(1+N))
dove$normal = qnorm(dove$rank/(1+N))

nbreak = max(cbind(length(seq(min(hawk$normal), max(hawk$normal), by=0.1)),
                   length(seq(min(jane$normal), max(jane$normal), by=0.1)),
                   length(seq(min(dove$normal), max(dove$normal), by=0.1))))
```

The standardized scores are then displayed on individual histograms and again on a combined histogram.
```{r}
hawk_hist2 = hist(hawk$normal, breaks = nbreak,
     col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
     xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Hawk Scores",
     xlab="Standard Deviations from Mean",
     ylab="Frequency")

jane_hist2 = hist(jane$normal, breaks = nbreak, 
     col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
     xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Jane Scores",
     xlab="Standard Deviations from Mean",
     ylab="Frequency")

dove_hist2 = hist(dove$normal, breaks = nbreak,
     col=rgb(0,0,255,a, maxColorValue = 255), lwd=2,
     xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Dove Scores",
     xlab="Standard Deviations from Mean",
     ylab="Frequency")

hawk_hist2 = hist(hawk$normal, breaks = nbreak,
     col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
     xlim=range(-2.5,2.5), ylim=c(0,ymax), main="All Standardized Scores",
     xlab="Standard Deviations from Mean",
     ylab="Frequency")

par(new=TRUE)

jane_hist2 = hist(jane$normal, breaks = nbreak, 
     col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
     xlim=range(-2.5,2.5), ylim=c(0,ymax), main=NULL, xlab=NULL, ylab=NULL)

par(new=TRUE)

dove_hist2 = hist(dove$normal, breaks = nbreak,
     col=rgb(0,0,255,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main=NULL, xlab=NULL, ylab=NULL)

scale_factor2 = N*0.3
all3_2 = sort(c(hawk$normal, jane$normal, dove$normal))
y2 = density(all3_2)
y_scaled2 = scale_factor2*y2$y
lines(y2$x, y_scaled2, col = "black", lwd = 2)
```

# Repeat the simulation with less applications

The number of applications is set to 20, and everything is recomputed.
```{r}
N = 20
```

```{r}
set.seed(1)
hawk1 = rsnorm(N, mean = 9, sd = 4, xi = 3)
hawk = data.frame(Raw_Hawk = as.vector(round(hawk1)))
hawk$Raw_Hawk_Sorted = sort(hawk$Raw_Hawk)
hawk_hist = hist(hawk$Raw_Hawk_Sorted, breaks = seq(min(hawk$Raw_Hawk)-1,
  max(hawk$Raw_Hawk)), plot = FALSE)
ymax = max(hawk_hist$counts)

hawk_hist = hist(hawk$Raw_Hawk_Sorted, breaks = seq(min(hawk$Raw_Hawk)-1,
  max(hawk$Raw_Hawk), by=1),  col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
  xlim=range(0,50), ylim=c(0,ymax), main="Hawk Scores - Raw", xlab="Score",
  ylab="Frequency")
```
```{r}
set.seed(1)
jane1 = rnorm(N, 19, 7)
jane = data.frame(Raw_Jane = as.vector(round(jane1)))
jane$Raw_Jane_Sorted = sort(jane$Raw_Jane)

jane_hist = hist(jane$Raw_Jane_Sorted, breaks = seq(min(jane$Raw_Jane)-1,
  max(jane$Raw_Jane), by=1), col=rgb(0,0,255,a, maxColorValue = 255), lwd=2, 
  xlim=range(0,50), ylim=c(0, ymax), main="Jane Scores - Raw", xlab="Score", 
  ylab="Frequency")
```
```{r}
set.seed(1)
dove1 = rsnorm(N, mean = 40, sd = 4, xi = 0.5)
dove = data.frame(Raw_Dove = as.vector(round(dove1)))
dove$Raw_Dove_Sorted = sort(dove$Raw_Dove)

dove_hist = hist(dove$Raw_Dove_Sorted, breaks = seq(min(dove$Raw_Dove)-1, 
  max(dove$Raw_Dove), by=1), col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
  xlim=range(0,50), ylim=c(0,ymax), main="Dove Scores - Raw", xlab="Score",
  ylab="Frequency")
```
```{r}
mydata = data.frame(App_No = c(1:N),
                    Hawk_raw = hawk$Raw_Hawk_Sorted,
                    Jane_raw = jane$Raw_Jane_Sorted,
                    Dove_raw = dove$Raw_Dove_Sorted
                                          )
head(mydata, n=20)

```
```{r}
ymax = max(c(hawk_hist$counts, jane_hist$counts, dove_hist$counts))

hist(hawk$Raw_Hawk_Sorted, breaks = seq(min(hawk$Raw_Hawk)-1,
                                        max(hawk$Raw_Hawk), by=1),
  col=rgb(255,0,0,a, maxColorValue = 255), lwd=2, xlim=range(0,50), 
  ylim=range(0,ymax), main=NULL, xlab=NULL, ylab=NULL)
par(new=TRUE)
hist(jane$Raw_Jane_Sorted, breaks = seq(min(jane$Raw_Jane)-1,
                                        max(jane$Raw_Jane), by=1),
  col=rgb(0,0,255,a, maxColorValue = 255), lwd=2, xlim=range(0,50),
  ylim=range(0,ymax), main=NULL, xlab=NULL, ylab=NULL)
par(new=TRUE)
hist(dove$Raw_Dove_Sorted, breaks = seq(min(dove$Raw_Dove)-1,
                                        max(dove$Raw_Dove), by=1),
  col=rgb(0,255,0,a, maxColorValue = 255), lwd=2, xlim=range(0,50),
  ylim=range(0,ymax), main="Raw Scores", xlab="Score", ylab="Frequency")

scale_factor = N*3
x = seq(0, 40, length.out = 3*N)
all3 = sort(c(hawk$Raw_Hawk, jane$Raw_Jane, dove$Raw_Dove))

y2 = density(all3)
y5 = scale_factor*y2$y
lines(y2$x, y5, col = "black", lwd = 2)
```
```{r}
hawk$rank = rank(hawk$Raw_Hawk)
jane$rank = rank(jane$Raw_Jane)
dove$rank = rank(dove$Raw_Dove)

hawk$normal = qnorm(hawk$rank/(1+N))
jane$normal = qnorm(jane$rank/(1+N))
dove$normal = qnorm(dove$rank/(1+N))

nbreak = max(cbind(length(seq(min(hawk$normal), max(hawk$normal), by=0.1)),
                   length(seq(min(jane$normal), max(jane$normal), by=0.1)),
                   length(seq(min(dove$normal), max(dove$normal), by=0.1))))
```
```{r}
hawk_hist2 = hist(hawk$normal, breaks = nbreak,
     col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Hawk Scores",
xlab="Standard Deviations from Mean",
ylab="Frequency")

jane_hist2 = hist(jane$normal, breaks = nbreak, 
     col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Jane Scores",
xlab="Standard Deviations from Mean",
ylab="Frequency")

dove_hist2 = hist(dove$normal, breaks = nbreak,
     col=rgb(0,0,255,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main="Standardized Dove Scores",
xlab="Standard Deviations from Mean",
ylab="Frequency")

hawk_hist2 = hist(hawk$normal, breaks = nbreak,
     col=rgb(255,0,0,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main="All Standardized Scores",
xlab="Standard Deviations from Mean",
ylab="Frequency")

par(new=TRUE)

jane_hist2 = hist(jane$normal, breaks = nbreak, 
     col=rgb(0,255,0,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main=NULL, xlab=NULL, ylab=NULL)

par(new=TRUE)

dove_hist2 = hist(dove$normal, breaks = nbreak,
     col=rgb(0,0,255,a, maxColorValue = 255), lwd=2,
xlim=range(-2.5,2.5), ylim=c(0,ymax), main=NULL, xlab=NULL, ylab=NULL)

scale_factor2 = N*0.3
all3_2 = sort(c(hawk$normal, jane$normal, dove$normal))
y2 = density(all3_2)
y_scaled2 = scale_factor2*y2$y
lines(y2$x, y_scaled2, col = "black", lwd = 2)
```



